Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-02 ms.qmd
Fist section of methods fixed. Up to Data Compilation

Reversible phenotypic plasticity, such as acclimation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses are relatively rapid and can therefore be fine-tuned to proximate environmental conditions [assuming the costs of plasticity are low – @Dewitt1998; @Scheiner2020]. Acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to faciliate a rapid response to the environment. However, the focus up to now has been primarily on mean physiological responses. For example, mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022]. However, it is possible that intrapopulation variability might also be impacted in addition to the mean. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance also suggests strong stabilising selection or reflect constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are, as of yet, poorly measured and modelled.

@@

We split the screening of titles and abstracts for the 1,321 papers found in our search among all authors evenly. To ensure consistency among authors in title and abstracts that should be included, prior to  screening all authors went through a randomly selected set of papers together - agreeing on those that were relevant and those that were not based on our inclusion criteria (see below). Where any authors were uncertain about whether to include a paper in the sub-sample they screened, we conservatively included the paper for full text screening and discussed uncertain papers among authors to come to a decision on whether to include the paper. After title and abstract screening, we were left with a total of 149 papers for full text screening. Papers were included only if they: 1) measured a physiological rate acutely at two temperatures on a sample of animals chronically exposed to the same two temperatures for at least 1 week; and 2) where physiological rates measured were burst and sustained locomotion, metabolic rates (standard, resting, routine and maximal), heart rates, and/or enzyme activities. We provide a PRISMA flow diagram of our extraction process in the *Supplement* (see @fig-prisma).

>>>>

Reversible phenotypic plasticity, such as acclimation, is expected to provide greater potential to buffer populations from climate impacts as responses are relatively rapid and can therefore be fine-tuned to proximate environmental conditions [assuming the costs of plasticity are low – @Dewitt1998; @Scheiner2020]. Acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to faciliate a rapid response to the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses. For example, mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022]. However, it is possible that intrapopulation variability might also be impacted in addition to the mean. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance also suggests strong stabilising selection or reflect constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms which may have important cascading effects on energy transfer and productivity across different ecosystems.

@@

We split the screening of titles and abstracts for the 1,321 papers found in our search among DWAN, FK, FS, and SN evenly. To ensure consistency among authors in title and abstract inclusion, relevant authors went through a randomly selected set of papers together before the formal screening to calibrate selection of papers based on our inclusion criteria (see below). In cases of disagreement regarding inclusion, we conservatively included the paper for full text screening and discussed uncertain papers among authors to come to a decision. After title and abstract screening, there was a total of 149 papers for full text screening. Papers were included only if they: 1) measured a physiological rate acutely at two temperatures on a sample of animals chronically exposed to the same two temperatures for at least 1 week; and 2) where physiological rates measured were burst and sustained locomotion, metabolic rates (standard, resting, routine and maximal), heart rates, and/or enzyme activities. We provide a PRISMA flow diagram of our extraction process in the *Supplement* (see @fig-prisma).
